Alluvial Diagrams in ggplot2

### 2018-02-11 inst/doc/ggalluvial.rmd

---
title: "alluvial diagrams in ggplot2"
author: "Jason Cory Brunson"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{alluvial diagrams in ggplot2}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

@@

Many other resources exist for visualizing categorical data in R, including several more basic plot types that are likely to more accurately convey proportions to viewers when the data are not so structured as to warrant an alluvial diagram. In particular, check out Michael Friendly's [**vcd** and **vcdExtra** packages (PDF)](https://cran.r-project.org/web/packages/vcdExtra/vignettes/vcd-tutorial.pdf) for a variety of statistically-motivated categorical data visualization techniques, Hadley Wickham's [**productplots** package](https://github.com/hadley/productplots) and Haley Jeppson and Heike Hofmann's descendant [**ggmosaic** package](https://cran.r-project.org/web/packages/ggmosaic/vignettes/ggmosaic.html) for product or mosaic plots, and Nicholas Hamilton's [**ggtern** package](http://www.ggtern.com/) for ternary coordinates. Other related packages are mentioned below.

@@

- An *axis* is a dimension (variable) along which the data are vertically grouped at a fixed horizontal position. The diagram above uses three categorical axes: `Class`, `Sex`, and `Age`.
- The groups at each axis are depicted as opaque blocks called *strata*. For example, the `Class` axis contains four strata: `1st`, `2nd`, `3rd`, and `Crew`.
- Horizontal (x-) splines called *alluvia* span the width of the diagram. In this diagram, each alluvium corresponds to a fixed value of each axis variable, indicated by its vertical position at the axis, as well as of the `Survived` variable, indicated by its fill color.
- The segments of the alluvia between pairs of adjacent axes are *flows*.
- The alluvia intersect the strata at *lodes*, The lodes are not visualized in the above diagram, but they can be inferred as filled rectangles extending the flows through the strata at each end of the diagram or connecting the flows on either side of the center stratum.

@@

**ggalluvial** recognizes two formats of "alluvial data", detailed in the following subsections. A third, tabular, form is popular for storing categorical data such as the `Titanic` and `UCBAdmissions` datasets used in this vignette.[^tableform] For consistency with tidy data principles and **ggplot2** conventions, **ggalluvial** does not accept tabular input.

@@

### Alluvia format

The first format reflects the visual arrangement of an alluvial diagram, but "untwisted": _Each row corresponds to a cohort of observations that take a specific value at each variable, and each variable has its own column._ An additional column contains the weight of each row, e.g. the number of observational units in the cohort. This is the format into which the base function `as.data.frame()` transforms a frequency table, for instance the 3-dimensional `UCBAdmissions` dataset:

@@

This format is inherited from the first version of **ggalluvial**, which modeled it after usage in **alluvial**. It required a stark departure from the usual position aesthetics: The user declares any number of axis variables, which `stat_alluvium()` and `stat_stratum()` recognize and process in a consistent way:

```{r alluvial diagram of UC Berkeley admissions dataset}
ggplot(as.data.frame(UCBAdmissions),
       aes(weight = Freq, axis1 = Gender, axis2 = Dept)) +
  geom_alluvium(aes(fill = Admit), width = 1/12) +
  geom_stratum(width = 1/12, fill = "black", color = "grey") +
  geom_label(stat = "stratum", label.strata = TRUE) +
  scale_x_continuous(breaks = 1:2, labels = c("Gender", "Dept")) +
  ggtitle("UC Berkeley admissions and rejections, by sex and department")
```

An important feature of these diagrams is the meaningfulness of the vertical axis: No gaps are inserted between the strata, so the total height of the diagram reflects the cumulative weight of the observations. The diagrams produced by **ggalluvial** conform (somewhat; see below) to the "grammar of graphics" principles of **ggplot2**, and this prevents users from producing "free-floating" diagrams like the Sankey diagrams showcased [here](https://developers.google.com/chart/interactive/docs/gallery/sankey).[^ggforce]
**ggalluvial** parameters and existing **ggplot2** functionality can also produce [parallel sets](https://eagereyes.org/parallel-sets) plots, illustrated here using the `Titanic` dataset:[^ggparallel]

@@

This format and functionality are useful and will be retained in future versions. They also involve some conspicuous deviations from **ggplot2** norms:

- The `axis[0-9]*` position aesthetics are non-standard.
- `stat_alluvium()` ignores any argument to the `group` aesthetic; instead, `StatAlluvium$compute_panel()` uses `group` to link the rows of the internally-transformed dataset that correspond to the same alluvium.
- The `label.strata` parameter instructs `stat_stratum()` (called by `geom_text()`) to take the values of the axis variables as labels.
- The horizontal axis must be manually corrected (using `scale_x_continuous()`) to reflect the implicit categorical variable identifying the axis.

@@

### Lodes format

The second format recognized by **ggalluvial** contains _one row per lode_, and can be understood as the result of "gathering" (in the **dplyr** sense) the axis columns of a dataset in the alluvia format into a key-value pair of columns encoding the axis as the `key` and the stratum as the `value`. This format requires an additional indexing `id` column that links the rows corresponding to a common cohort, i.e. the lodes of a single alluvium, as illustrated below using the `to_lodes()` defaults on the Berkeley admissions dataset:

@@

The same stat and geom can receive this data format using a different set of positional aesthetics, also specific to **ggalluvial**:

@@

Weights (and weight totals) can vary from axis to axis, allowing users to produce bump charts like those showcased [here](http://imgur.com/gallery/gI5p7).[^geom-area] In these cases, the strata are an artifact of the alluvia and usually not plotted. For convenience, both `stat_alluvium()` and `stat_flow()` accept arguments for `x` and `alluvium` even if none is given for `stratum`.[^arguments] As an example, we can grop countries in the `Refugees` dataset by region, in order to compare refugee volumes at different scales:

[^geom-area]: If bumping is unnecessary, consider using [`geom_area()`](https://chrisalbon.com/r-stats/stacked-area-graph.html) instead.
[^arguments]: `stat_stratum()` will similarly accept arguments for `x` and `stratum` without `alluvium`. If both strata and either alluvia or flows are to be plotted, though, all three parameters need arguments.

@@

```{r alluvial diagram of majors dataset}
data(majors)
majors$curriculum <- as.factor(majors$curriculum)
ggplot(majors,
       aes(x = semester, stratum = curriculum, alluvium = student,
           fill = curriculum, label = curriculum)) +
  geom_flow(stat = "alluvium", lode.guidance = "rightleft",
            color = "darkgray") +
  geom_stratum() +
  theme(legend.position = "bottom") +
  ggtitle("student curricula across several semesters")
```

No `weight` is specified, so each row is assigned a weight of `1`.
This example demonstrates one way **ggalluvial** handles missing data. The alternative is to set the parameter `na.rm` to `TRUE`.[^na.rm] Missing data handling (specifically, the order of the strata) also depends on whether the `stratum` variable is character or factor/numeric.

>>>>

---
title: "Alluvial Diagrams in ggplot2"
author: "Jason Cory Brunson"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{alluvial diagrams in ggplot2}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

@@

Many other resources exist for visualizing categorical data in R, including several more basic plot types that are likely to more accurately convey proportions to viewers when the data are not so structured as to warrant an alluvial diagram. In particular, check out Michael Friendly's [**vcd** and **vcdExtra** packages (PDF)](https://CRAN.R-project.org/package=vcdExtra/vignettes/vcd-tutorial.pdf) for a variety of statistically-motivated categorical data visualization techniques, Hadley Wickham's [**productplots** package](https://github.com/hadley/productplots) and Haley Jeppson and Heike Hofmann's descendant [**ggmosaic** package](https://CRAN.R-project.org/package=ggmosaic/vignettes/ggmosaic.html) for product or mosaic plots, and Nicholas Hamilton's [**ggtern** package](http://www.ggtern.com/) for ternary coordinates. Other related packages are mentioned below.

@@

- An *axis* is a dimension (variable) along which the data are vertically grouped at a fixed horizontal position. The diagram above uses three categorical axes: `Class`, `Sex`, and `Age`.
- The groups at each axis are depicted as opaque blocks called *strata*. For example, the `Class` axis contains four strata: `1st`, `2nd`, `3rd`, and `Crew`.
- Horizontal (x-) splines called *alluvia* span the width of the diagram. In this diagram, each alluvium corresponds to a fixed value of each axis variable, indicated by its vertical position at the axis, as well as of the `Survived` variable, indicated by its fill color.
- The segments of the alluvia between pairs of adjacent axes are *flows*.
- The alluvia intersect the strata at *lodes*. The lodes are not visualized in the above diagram, but they can be inferred as filled rectangles extending the flows through the strata at each end of the diagram or connecting the flows on either side of the center stratum.

@@

**ggalluvial** recognizes two formats of "alluvial data", treated in detail in the following subsections, but which basically correspond to the "wide" and "long" formats of categorical repeated measures data. A third, tabular (or array), form is popular for storing data with multiple categorical dimensions, such as the `Titanic` and `UCBAdmissions` datasets.[^tableform] For consistency with tidy data principles and **ggplot2** conventions, **ggalluvial** does not accept tabular input; `base::data.frame()` converts such an array to an acceptable data frame.

@@

### Alluvia (wide) format

The wide format reflects the visual arrangement of an alluvial diagram, but "untwisted": _Each row corresponds to a cohort of observations that take a specific value at each variable, and each variable has its own column._ An additional column contains the weight of each row, e.g. the number of observational units in the cohort. This is the format into which the base function `as.data.frame()` transforms a frequency table, for instance the 3-dimensional `UCBAdmissions` dataset:

@@

This format is inherited from the first version of **ggalluvial**, which modeled it after usage in **alluvial**: The user declares any number of axis variables, which `stat_alluvium()` and `stat_stratum()` recognize and process in a consistent way:

```{r alluvial diagram of UC Berkeley admissions dataset}
ggplot(as.data.frame(UCBAdmissions),
       aes(weight = Freq, axis1 = Gender, axis2 = Dept)) +
  geom_alluvium(aes(fill = Admit), width = 1/12) +
  geom_stratum(width = 1/12, fill = "black", color = "grey") +
  geom_label(stat = "stratum", label.strata = TRUE) +
  scale_x_discrete(limits = c("Gender", "Dept"), expand = c(.05, .05)) +
  scale_fill_brewer(type = "qual", palette = "Set1") +
  ggtitle("UC Berkeley admissions and rejections, by sex and department")
```

An important feature of these diagrams is the meaningfulness of the vertical axis: No gaps are inserted between the strata, so the total height of the diagram reflects the cumulative weight of the observations. The diagrams produced by **ggalluvial** conform (somewhat; keep reading) to the "grammar of graphics" principles of **ggplot2**, and this prevents users from producing "free-floating" diagrams like the Sankey diagrams showcased [here](https://developers.google.com/chart/interactive/docs/gallery/sankey).[^ggforce]
**ggalluvial** parameters and existing **ggplot2** functionality can also produce [parallel sets](https://eagereyes.org/parallel-sets) plots, illustrated here using the `Titanic` dataset:[^ggparallel]

@@

This format and functionality are useful for many applications and will be retained in future versions. They also involve some conspicuous deviations from **ggplot2** norms:

- The `axis[0-9]*` position aesthetics are non-standard: they are not an explicit set of parameters but a family based on a regular expression pattern; and at least one, but no specific one, is required.
- `stat_alluvium()` ignores any argument to the `group` aesthetic; instead, `StatAlluvium$compute_panel()` uses `group` to link the rows of the internally-transformed dataset that correspond to the same alluvium.
- The `label.strata` parameter instructs `stat_stratum()` (called by `geom_text()`) to take the values of the axis variables as labels.
- The horizontal axis must be manually corrected (using `scale_x_discrete()` or `scale_x_continuous()`) to reflect the implicit categorical variable identifying the axis.

@@

### Lodes (long) format

The long format recognized by **ggalluvial** contains _one row per lode_, and can be understood as the result of "gathering" (in the **dplyr** sense) or "pivoting" (in the Microsoft Excel sense) the axis columns of a dataset in the alluvia format into a key-value pair of columns encoding the axis as the key and the stratum as the value. This format requires an additional indexing column that links the rows corresponding to a common cohort, i.e. the lodes of a single alluvium, as illustrated below using the `to_lodes()` defaults on the Berkeley admissions dataset:

@@

The functions that convert data between wide (alluvia) and long (lodes) format include several parameters that help preserve ancillary information. See `help("alluvial-data")` for examples.

The same stat and geom can receive data in this format using a different set of positional aesthetics, also specific to **ggalluvial**:

@@

Weights (and weight totals) can vary from axis to axis, allowing users to produce bump charts like those showcased [here](http://imgur.com/gallery/gI5p7).[^geom-area] In these cases, the strata are an artifact of the alluvia and often not plotted. For convenience, both `stat_alluvium()` and `stat_flow()` will accept arguments for `x` and `alluvium` even if none is given for `stratum`.[^arguments] As an example, we can group countries in the `Refugees` dataset by region, in order to compare refugee volumes at different scales:

[^geom-area]: If bumping is unnecessary, consider using [`geom_area()`](http://www.r-graph-gallery.com/136-stacked-area-chart/) instead.
[^arguments]: `stat_stratum()` will similarly accept arguments for `x` and `stratum` without `alluvium`. If both strata and either alluvia or flows are to be plotted, though, all three parameters need arguments.

@@

```{r alluvial diagram of majors dataset}
data(majors)
majors$curriculum <- as.factor(majors$curriculum)
ggplot(majors,
       aes(x = semester, stratum = curriculum, alluvium = student,
           fill = curriculum, label = curriculum)) +
  scale_fill_brewer(type = "qual", palette = "Set2") +
  geom_flow(stat = "alluvium", lode.guidance = "rightleft",
            color = "darkgray") +
  geom_stratum() +
  theme(legend.position = "bottom") +
  ggtitle("student curricula across several semesters")
```

No `weight` is specified, so each row is assigned unit weight.
This example demonstrates one way **ggalluvial** handles missing data. The alternative is to set the parameter `na.rm` to `TRUE`.[^na.rm] Missing data handling (specifically, the order of the strata) also depends on whether the `stratum` variable is character or factor/numeric.
